---
title: "Assignment 4 - Mari Herbst de Cortina"
author: "Mari Herbst"
date: "11/3/2021"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(here)
library(lubridate)
library(kableExtra)
library(ggbeeswarm)
library(effsize)
library(modelsummary)
```

# Introduction

This report explores data collected about snowshoe hares across sites in the Bonanza Creek LTER between 1998 and 2012 (Kielland et al. (2017). The report investigates the data relating to juvenile hares specifically, plotting the number of juvenile hares recorded each year and the distribution in weight vs. sex at each grid site. It also explores whether there is a linear relationship between juvenile hare weights and hind foot length.

# Data and analysis

Snowshoe hares in the Bonanza Creek LTER in Alaska were trapped and measured over the course of 14 years (1998-2012). The hares were observed at three locations -- the Bonanza Riparian zone, the Bonanza Black Spruce zone, and the Bonanza Mature zone, representing three different landscapes in the region. The hares were measured for sex (male, female, NA), age (adult, juvenile, NA), weight (g), and hind foot length (mm).

This report visualizes juvenile hares observed each year, the distribution of weights for juvenile hares according to their sex and collection site, and the relationship between weight and hind foot length in juvenile hares. Statistical significance of the weight differences between sexes was measured with a two-sample t-test using a significance level of 0.05. Cohen's d describes the effect size, or the difference between the two groups. Finally, a simple linear regression is used to explore the potential relationship between hind foot length and weight. All analyses are in R version 4.1.1 using RStudio version 1.4.1717.

data citation: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14


```{r, message = FALSE, warning = FALSE}
# Read in the data using here() to create a system agnostic file path.
bonanza_hares <- read_csv(here("data", "bonanza_hares.csv"))
```

# Juvenile hare investigations

## Annual juvenile hare trap counts

```{r, message = FALSE, warning = FALSE}
# Transform the data to separate year into its own column
hares_year <- bonanza_hares %>%
  mutate(date = mdy(date)) %>% # change class to date
  mutate(year = year(date))
```


```{r, message = FALSE, warning = FALSE}
# Summarize the data to get a count of trapped juvenile hares per year
hares_summary <- hares_year %>%
  filter(age == "j") %>% # select only juvenile hare observations
  group_by(year) %>% 
  summarize(hare_count = n()) # summarize hares_year by year and count of hares trapped
```

```{r, message = FALSE, warning = FALSE}
# Create a plot of trapped hare counts per year
ggplot(data = hares_summary, aes(x = year, y = hare_count)) + # create a plot of hare trapped counts per year
  geom_col(fill = "darkolivegreen") + # in a bar graph where the bars are colored olive green
  scale_x_continuous(breaks = 1998:2012) + # range of years
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), # rotate x-axis labels for easier reading
        text = element_text(family = "Calibri", # change font and make labels bold
                            face = "bold"),
        plot.title = element_text(hjust = 0.5)) + # center plot title
  labs(x = "year", # label axes
       y = "count",
       title = "Count of juvenile hares trapped per year") # add graph title
```

Figure 1. Count of hares trapped per year: Snowshoe hares were trapped and released over the course of 1998-2012 at three locations in the Tanana Valley in Alaska.

### Major takeaways

The most juvenile hares recorded in a single year was `r max(hares_summary$hare_count)`. No juvenile hares were trapped in 1998, 2002, nor 2009. The mean number of juvenile snowshoe hare observations from 1998-2012 is `r sum(hares_summary$hare_count)/12` and the median value is `r median(hares_summary$hare_count)`. The trends in hare counts displayed on this graph show a decline in juvenile hare counts over time with occasional resurgances, though this may be due to irregularity in collection. To standardize juvenile hare populations for future analyses, I would recommend creating a more regular interval for hare data collection. The dates of hare collections are varied over the course of the years where data is presented. They frequently represent different times of year where hare populations may differ for various reasons that are not isolated by the nature of the study set up at present. Additionally, there are a different number of days when observations were taken across the different years which likely influenced the number of hares recorded.

## Visualize juvenile hare weights

```{r, message = FALSE, warning = FALSE}
# Change labels to make data easier to understand once visualized
bonanza_clean <- bonanza_hares %>%
  filter(age == "j") %>% # filter for juvenile hares
  mutate(sex = case_when( # when sex =
         sex == "m" ~ "male", # "m" change to "male"
         sex == "f" ~ "female")) %>%
  mutate(grid = case_when( # when grid =
         grid == "bonbs" ~ "Bonanza Black Spruce", # "bonbs" change to "Bonanza Black Spruce"
         grid == "bonrip" ~ "Bonanza Riparian",
         grid == "bonmat" ~ "Bonanza Mature"))
```

```{r, message = FALSE, warning = FALSE}
# Create a plot of juvenile hare weights by sex and collection site (grid)
ggplot(data = bonanza_clean, aes(x = sex, y = weight)) + # only need 1 variable for a histogram, the other is count
  geom_beeswarm(na.rm = FALSE, aes(color = sex)) + # include NA values, assign color depending on sex
  facet_wrap(~ grid) + # divide data by collection site
  geom_boxplot(fill = NA, outlier.colour = NA) + # add a boxplot to visualize summary statistics
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), # rotate x axis values for easier reading
        text = element_text(family = "Calibri", # change font and make labels bold
                            face = "bold"),
        plot.title = element_text(hjust = 0.5)) +  # center plot title
  labs(x = "sex", y = "weight (g)", title = "Juvenile Hare Weight Distribution by Sex and Grid Site") # label axes and title
```

Figure 2. Juvenile Hare Weight Distributions by Sex and Grid Site: Snowshoe hare weight distributions are examined across the Bonanza Black Spruce, Bonanza Mature, and Bonanza Riparian data collection sites. NA indicates the sex was not identified. The middle line of the box plot represents the median value of the data set, the distances between the median line and the upper and lower box boundaries indicate the middle number between the median and further extreme value (highest or lowest, respectively). The whiskers extend form the box boundaries (explained above) to the most extreme value in either direction.

## Juvenile weight comparison by sex

```{r, message = FALSE, warning = FALSE}

# create a data frame for only juvenile males and females
hares_j_m_f <- bonanza_clean %>% # start with bonanza_clean
  filter(sex == "male" | sex == "female") # filter for observations where sex is male or female

# Create a table of summary statistics for hare weights
weight_summary_stats <- hares_j_m_f %>%
  group_by(sex) %>%
  summarize(mean_weight = round(mean(weight, na.rm = TRUE), 2), # 'na.rm = TRUE' to remove NA values, round() to choose number of digits
            median_weight = median(weight, na.rm = TRUE), 
            sd_weight = round(sd(weight, na.rm = TRUE), 2),
            sample_size = n())

# Finalize table
weight_summary_stats %>%
  kable(col.names = c("sex", "mean weight (g)", "median weight (g)", "standard deviation (g)", "sample size" ), # finalized column names
        caption = "Table 1. Summary statistics for juvenile snowshoe hares trapped and recorded between 1998-2012 in the Bonanza Creek LTER") %>%
  kable_paper() #styling
```

```{r, include = FALSE, message = FALSE, warning = FALSE}
# exploratory visualizations

# histogram to check for normal distribution
ggplot(data = hares_j_m_f, aes(x = weight)) +
  geom_histogram(aes(fill = sex)) +
  facet_wrap(~ sex)

# q-q plot to check for normal distribution
ggplot(data = hares_j_m_f, aes(sample = weight)) +
  stat_qq() +
  facet_wrap(~ sex) +
  theme_minimal()
```

### t-test and major takeaways

```{r, message = FALSE, warning = FALSE}
female_hares <- hares_j_m_f %>% # create a vector for just female hare weights
  filter(sex == "female") %>% # filter for female only
  pull(weight) # select only the weight column values

male_hares <- hares_j_m_f %>% # create a vector for just male hare weights
  filter(sex == "male") %>% # filter for male only
  pull(weight) # select only the weight column values

# means and SD of each sex
mean_female <- mean(female_hares, na.rm = TRUE)
mean_male <- mean(male_hares, na.rm = TRUE)
sd_female <- sd(female_hares, na.rm = TRUE)
sd_male <- sd(female_hares, na.rm = TRUE)

# t-test
hare_ttest <- t.test(female_hares, male_hares)

# find effect size
hare_effsize <- cohen.d(female_hares, male_hares, na.rm = TRUE)
```

On average, male snowshoe hares weigh more than female snowshoe hares, `r round(mean_male, 2)`g $\pm$ `r round(sd_male, 2)`g and `r round(mean_female, 2)`g $\pm$ `r round(sd_female, 2)`g, respectively (mean (g) $\pm$ one standard deviation (g)). The absolute difference in means is `r round((mean_male-mean_female), 2)`g, which amounts to a `r round((mean_male - mean_female)/((mean_male + mean_female)/2) * 100, 2)`% difference in means. This difference in means is significant (Welch's Two Sample t-test: t(`r round(hare_ttest$parameter, 2)`) = `r round(hare_ttest$statistic, 2)`, p = `r hare_ttest$p.value`), but the effect size, `r round(hare_effsize$estimate, 2)`, is small.

## Hare weight and hind foot length

```{r, message = FALSE, warning = FALSE}
ggplot(data = bonanza_clean, aes(x = hindft, y = weight)) +
  geom_point(color = "deeppink3") +
  theme_minimal() +
  theme(text = element_text(family = "Calibri", # change the font and bold for easier reading
                            face = "bold"),
        plot.title = element_text(hjust = 0.5)) +  # center plot title
  labs(x = "hind foot length (mm)",
       y ="weight (g)",
       title = "Juvenile Hare Weight by Hind Foot Length") +
  geom_smooth(method = "lm", color = "lightslateblue")
```

Figure 3. Juvenile Hare Weight by Hind Foot Length: An exploration of the trends of `r sum(weight_summary_stats$sample_size)` snowshoe hares' weights (g) relative to their hind foot lengths (mm).

```{r, include = FALSE, message = FALSE, warning = FALSE}
# explore the linearity of hare weights/hind feet length relationship
hare_model <- lm(weight ~ hindft, data = bonanza_clean)

# call the model:
hare_model

# get more information about the model
summary(hare_model)

# create a tidy table of the model
hare_model_coeff <- tidy(hare_model)

# explore the coefficient of correlation for hare weight and hind foot length, Pearson's r
hare_pearsonr <- cor.test(bonanza_clean$weight, bonanza_clean$hindft)

# explore scedasticity
plot(hare_model)
```

### Major takeaways

Based on this model, we would expect an average increase in hare body weight of `r round(hare_model_coeff$estimate[2], 2)`g per 1mm increase in hind foot length. The model also indicates that hind foot length is predictive of `r round(summary(hare_model)$r.squared * 100, 2)`% of variance in snowshoe hare weights. There is moderate positive correlation between hare hind foot length and weight (Pearson's r: `r round(hare_pearsonr$estimate, 2)`). I have some concerns about making the assumption that this data set can be described by a linear regression. The R$^2$ value and Pearson's r demonstrate that there is some correlation between hind foot length and weight, but not a strong correlation enough correlation to explain all variance in weights. Further, the data appears to demonstrate heteroscedasticity rather than homoscedasticity, indicating there is not constant variance in the errors of observations.

# Summary

- The number of juvenile snowshoe hares trapped between 1998 and 2012 in the Tonata Valley of Alaska varied widely, potentially due to inconsistency in time of year when data was collected.

- Juvenile hare weights vary based on collection site and gender. The difference in means between males and females is significant, but the effect size is small.

- Juvenile hare weight and hind foot length are loosely correlated, but there is not enough correlation to explain the full variance in weights. Also, there are concerns about using a linear regression model for this data because the assumptions for linear regression are not met.

# Citations

## Data citation

Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14

## Package citations

Erik Clarke and Scott Sherrill-Mix (2017). ggbeeswarm: Categorical Scatter (Violin Point) Plots.
  R package version 0.6.0. https://CRAN.R-project.org/package=ggbeeswarm
  
Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with lubridate. Journal of
  Statistical Software, 40(3), 1-25. URL https://www.jstatsoft.org/v40/i03/.
  
Hao Zhu (2021). kableExtra: Construct Complex Table with 'kable' and Pipe Syntax. R package
  version 1.3.4. https://CRAN.R-project.org/package=kableExtra
  
Kirill Müller (2020). here: A Simpler Way to Find Your Files. R package version 1.0.1.
  https://CRAN.R-project.org/package=here

Torchiano M (2020). _effsize: Efficient Effect Size Computation_. doi: 10.5281/zenodo.1480624
  (URL: https://doi.org/10.5281/zenodo.1480624), R package version 0.8.1, <URL:
  https://CRAN.R-project.org/package=effsize>.
  
Vincent Arel-Bundock (2021). modelsummary: Summary Tables and Plots for Statistical Models and
  Data: Beautiful, Customizable, and Publication-Ready. R package version 0.9.4.
  https://CRAN.R-project.org/package=modelsummary

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686,
  https://doi.org/10.21105/joss.01686